When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-12-28 doc/paper/vpd_et_paper.tex
wrote through conlcusions, now need to go back and just read throuhg (improving plots), then can send out to pierre.

The idealized representation of ET used here is successful in developing intuition for how ET responds to changes in $VPD$. This intuition will aid the community in interpreting observations and output from sophisticated full complexity climate models.

The idealized framework leads to the following general conclusions:
\begin{itemize}
  \item Aerodynamic resistance plays an important role of scaling $\frac{\partial \; ET}{\partial \; VPD}$. This is a leading order effect for observing higher magnitude responses in DBF and ENF.
  \item In general, CSH has the most negative (i.e. ET reduced) response to increases in $VPD$ (atmospheric drying). So CSH plants will almost always try and conserve water, effectively reducing ET with dry atmospheric perturbation.
    \item Additionally for CSH, $VPD$ variability contributes the most to $ET$ variability.
\item CRO has the most positive response (i.e. ET increased) in response to increases in $VPD$. This is consistent with CROs that may be evolved or bred to thrive in non-water-limited environments.
\item The response is more a function of the environment for DBF, ENF, and GRA. Because as VPD increases the response is more likely to be positive, if RH is fixed then the response will be more likely to be positive at warmer T, or if T is fixed the response is more likely to be positive with decreasing RH.
\item ENF has the strongest dependence on environmental conditions due to its small $g1$.
\item Model and observational uncertainty is highest for GRA and CRO, so conclusions about those PFTs should be tempered.
\item However, inclusion of uncertainty doesn't alter conclusions about DBF, ENF, and CSH.
\end{itemize}

The intuition developed using this framework can be used to understand how the land surface will respond and contribute to changes in the environment. Additionally, Equation \ref{et} gives provides an estimate of ET that requires no additional information beyond that required to calculate PET. Given that for all PFTs, with the exception of CRO, we found a high frequency and magnitude of negative $\frac{\partial \; ET}{\partial \; VPD}$, PET is a physically unrealisitic representation of ET for vegetated surfaces and PET-based drought metrics are not usefull. We advocate for drought metrics using Equation \ref{et} instead of PET.

>>>>

We derived a new form of Penman Monteith using $uWUE$ (\cite{Zhou_2015}) to remove implicit depedence on ET in the stomatal conductance term. With our new form of Penman Monteith we deveoped a theory for when an ecosystem will tend to reduce or increase ET with increasing VPD. The goal was to capture the leading order behavior of the stem to gain some intrinsic knowledge for its behaviro. This intuition can be used to disentangle land atmosphere feedbacks in more complicated scenariors, and will aid interpretation of obserations and sophisticated models.

Our theory states that each plant type will have it's own critical VPD below which ET will decrease (plant response dominating) and above which ET will increase (atmospheric demand dominates). We tested our theory using data from FLUXNET 2015, and found that for DBF and CSH the theory succeeds in capturing the leading order behavor. For ENF, the thoery succeeds in capturing the leading order behavior for the sign of $\frac{\partial \; ET}{\partial \; VPD}$, but not for magnitude. In contrast, the model does not capture the behavior for CRO and GRA. In some ways we should have expected this, as our theory does not account for large variability we would expect for those plant types, including varying surface roughness, c3 vs c4 photosynthesis, and potential differences in irrigationa cross sites. Given these diffincies for CRO and GRA, we will withhold conclusions on their response to increasing VPD.

For the forest sites (DBF, ENF) the observed environmental VPD approximately straddles the critical VPD, with about half the observations below the critical VPD and half above. However for CSH the environmental VPD never exceeds the critical VPD, and even when uncertainty is included the sign of the derivative is almost always negative (93\% of observations).

While our theory predicts common occurence of positive $\frac{\partial \; ET}{\partial \; VPD}$ (atmospheric demand dominates), the plant response in attenuating ET from the land surface is still important even when $\frac{\partial \; ET}{\partial \; VPD} > 0$. The magnitude of the ``sign term'' of our theory is always far below what it would be if we only considered the atmospheric demand term. This means that any drought indicies or analysis using PET will not capture how the vegetated surface respond and react to increases in atmospheric demand. However, the new uWUE-version of Penman Monteith we derived could be used as a complement to PET in drought indices. Basically one could compare calculate and compare indices  when plant response is ignored (PET) or included (uWUE-Penman Monteith).

This paper provides the initial intuition for how the land surface respondes to atmospheric drying. This intuition for the one-way behavior of  land surface resonse to atmospheric drying is critical first step to disentangling land-atmosphere feedbacks at various scales, from diurnal to seasonal and beyond.
